---
title: "Documentation"
output: html_notebook
---

MSSL follows Hadley Wickham's guide to documenting functions as described at http://r-pkgs.had.co.nz/man.html:


#### Documentation Steps

1. Add roxygen comments to your .R files.

2. Run press Ctrl/Cmd + Shift + D in RStudio (or run devtools::document()) to convert roxygen comments to .Rd files. (devtools::document() calls roxygen2::roxygenise() to do the hard work.)

3. Preview documentation with ?.

4. Rinse and repeat until the documentation looks the way you want.


#### roxygen template

The simplest documentation should include the following:


```{r}
#' (Title of the documentation)
#'
#' (Description briefly describing what the function does)
#'
#' @param (parameter1 name) (description of parameter 1)
#' @param (parameter2 name) (description of parameter 2)
#' @param (parameter3 name) (description of parameter 3)
#'
#' @return (output name) (description of output) 
#'
#' @examples
#' (executable code that uses the function) 
#' 
```



#### roxygen sample 

This sample is straight from Hadley's guide. It shows code for the function sum(..., na.rm = FALSE). Note the use of formatting commands like \code and \link.

```{r}
#' Sum of vector elements.
#'
#' \code{sum} returns the sum of all the values present in its arguments.
#'
#' This is a generic function: methods can be defined for it directly
#' or via the \code{\link{Summary}} group generic. For this to work properly,
#' the arguments \code{...} should be unnamed, and dispatch is on the
#' first argument.
#'
#' @param ... Numeric, complex, or logical vectors.
#' @param na.rm A logical scalar. Should missing values (including NaN)
#'   be removed?
#' @return If all inputs are integer and logical, then the output
#'   will be an integer. If integer overflow
#'   \url{http://en.wikipedia.org/wiki/Integer_overflow} occurs, the output
#'   will be NA with a warning. Otherwise it will be a length-one numeric or
#'   complex vector.
#'
#'   Zero-length vectors have sum 0 by definition. See
#'   \url{http://en.wikipedia.org/wiki/Empty_sum} for more details.
#' @examples
#' sum(1:10)
#' sum(1:5, 6:10)
#' sum(F, F, F, T, T)
#'
#' sum(.Machine$integer.max, 1L)
#' sum(.Machine$integer.max, 1)
#'
#' \dontrun{
#' sum("a")
#' }
```



