---
title: "Ontologies in R"
author: "Thomas Brailey"
date: "May 2020"
output:
  html_document:
    df_print: paged
header-includes:
- \setlength\parindent{24pt}\setlength{\parskip}{0.0pt plus 1.0pt}
- \usepackage{setspace}\onehalfspacing
subtitle: Center for Peace and Security Studies (cPASS)
fontsize: 10pt
---

When dealing with hierarchical data, ontologies, dendrograms, and trees can be a good way of visualizing relationships between the so-called "parent" and "child" nodes of a dataset. This document provides a cursory overview of hierarchical data and some useful packages in R to deal with such data.

# Prep
```{r}
rm(list = ls())

`%>%` <- 
  magrittr::`%>%`
```

# Load and clean data
## Load
We start by loading the cleaned IISS data. This dataframe contains information about the count of military units at the country-year level. It is hierarchical in the sense that we break up military equipment into type and subtype. 
```{r}
# Load data
df <- 
  readRDS(file = "../../IISS/data/01e_addrows.rds")

df <- 
  data.frame(df)
```

## Prep
We need the data in tree format.
```{r}
# Temp recode carriers
df$tek[df$tek == "principal surface combatants_aircraft carriers_nuclear"] <- 
  "principal surface combatants_nuclear carriers"
df$tek[df$tek == "principal surface combatants_aircraft carriers_non-nuclear"] <- 
  "principal surface combatants_conventional carriers"

df <- 
  df %>% 
  dplyr::select(tek) %>%
  tidyr::separate(tek, c("parent", "child"), sep = "_", remove = TRUE) %>%
  dplyr::distinct() %>%
  dplyr::filter_all(dplyr::any_vars(!is.na(.)))

df$child[df$child == "all"] <- NA
```

# Create tree
Our first example uses the package *collapsibleTree*. 
```{r, fig.height=10, fig.width=15, fig.align="center"}
collapsible_tree <- collapsibleTree::collapsibleTree(df, hierarchy = c("parent", "child"), collapse = FALSE)

collapsible_tree
```

Our next example uses the package *data.tree*. Using *data.tree* requires a little bit of re-formatting, but nothing too complicated. Calling print() on the object df_tree shows us the new structure of the data frame. In order to plot the tree, simply call plot().
```{r}
# Remove NAs 
df$child[is.na(df$child)] <- ""
 
# Generate pathString as new column
df$pathString <- 
  paste("Tek",
        df$parent,
        df$child,
        sep = "|")

                             

# Create list 
df_tree <- 
  data.tree::as.Node(df, pathDelimiter = "|")

print(df_tree, limit = 15)
```

```{r}
plot(df_tree)
```

This isn't particularly helpful as a visualization, so we turn to the package *networkD3* to better visualize our data.tree object. 
```{r}
df_list <- 
  data.tree::ToListExplicit(df_tree, unname = TRUE)

df_vis <- 
  networkD3::radialNetwork(df_list)

df_vis
```

networkD3 has plenty of tree and dendrogram templates to choose from. 
```{r}
df_vis <- 
  networkD3::diagonalNetwork(df_list)

df_vis
```

For further reading, see the vignettes for the packages discussed above: 

 - [collapsibleTree](https://cran.r-project.org/web/packages/collapsibleTree/collapsibleTree.pdf) 

 - [data.tree](https://cran.r-project.org/web/packages/data.tree/vignettes/data.tree.html)

 - [networkD3](https://cran.r-project.org/web/packages/networkD3/networkD3.pdf)

There are many other packages in R designed to visualize heirarchical data. They all essentially have the same capabilities as the packages we have looked at in this document. Here are a few of them:

 - [igraph](https://igraph.org/r/)

 - [ontologyX](https://cran.r-project.org/web/packages/ontologyIndex/vignettes/intro-to-ontologyX.html)

 - [ggdendro](https://cran.r-project.org/web/packages/ggdendro/index.html)
 