---
title: "Waffle plots and heat plots in R"
author: "Thomas Brailey"
date: "May 2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \setlength\parindent{24pt}\setlength{\parskip}{0.0pt plus 1.0pt}
- \usepackage{setspace}\onehalfspacing
subtitle: Center for Peace and Security Studies (cPASS)
fontsize: 10pt
---

This document provides a cursory overview of waffle plots and heat plots in R. These plots can be useful for plotting large panel datasets. 

# Prep
Start by clearing your workspace and loading necessary packages (not included). 
```{r, include="false"}
rm(list = ls())

library(ggplot2)
`%>%` <- 
  magrittr::`%>%`
```

# Load and clean data
## Load
We start by loading a dataset on country military specialization scores. The unit of analysis is country-year-specialization score (not included).
```{r, include="false"}
# Load data
df <- 
  readRDS(file = "../../IISS/data/02a_DV_specialization.rds")

df <- 
  data.frame(df)
```

## Prep
After general cleaning, there isn't much in the way of data preparation. Just make sure that, if your highest level of analysis is a character variable, coerce it into a factor to allow for alphabetical reordering. 
```{r}
# Prep data
scores_waffle <- 
  df %>% 
  dplyr::filter(year != "   0") %>% 
  dplyr::mutate(country = forcats::fct_reorder(country, dplyr::desc(country)))

# Specify color palette
my_palette <- 
  RColorBrewer::brewer.pal(7, "Spectral")
```


# Plot
Waffle plots can be made using just ggplot2.  
```{r,  fig.width = 12, fig.height = 16}
p2 <- 
  ggplot(scores_waffle, aes(x = country, y = year, fill = d)) +
  geom_tile(color = "white", size = 0.35) +
  scale_fill_gradientn(colors = my_palette, na.value = 'white') +
  theme_bw() +
  coord_flip() +
  labs(title = "Military Specialization by Country by Year",
       x = "Country",
       y = "Year",
       fill = "Specialization Score",
       caption = "") +
  theme(panel.grid = element_blank(),
        title = element_text(size = 16),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.text.x = element_text(size = 12, angle = 90),
        axis.text.y = element_text(size = 8),
        strip.text = element_text(size = 12),
        legend.position = "bottom",
        legend.direction = "horizontal",
        legend.title = element_text())
p2
```

Voila! These types of plots can be customized in all kinds of ways. See below for expanding your heat plot knowledge: 

 - [Waffle and swarm plots](https://paulvanderlaken.com/2019/01/25/visualization-innovation-waffleplots-and-swarmplots-aka-beeplots/)
 
 - [Heat maps](https://www.r-graph-gallery.com/heatmap.html)
 
 - [Animated and interactive heat plots](https://www.r-bloggers.com/the-spread-of-covid-19-across-countries-visualization-with-r/)